---
title: "Read and format duckweed-bacteria coculture frond area data"
subtitle: ""
author: "Shane Hogle"
date: today
abstract: "This reads plate-reader exports of *Lemna* frond area measured at day 0, 4, and 8 of coculture with individual bacterial strains (or CONTROL, no bacteria), combines them into a single tibble, checks the plate/well design, and writes a cleaned table for downstream modeling."
---

*Lemna* was grown in 24-well plates, either alone (`CONTROL`) or co-cultured with a single bacterial strain per well group. Each plate hosts up to three strains + plant, with each strain occupying 8 wells (biological replicates) on that plate. The same physical plate was imaged at day 0, day 4, and day 8, so a given well is tracked longitudinally across the three timepoints.

## Libraries

```{r}
library(tidyverse)
library(here)
library(fs)
library(readxl)
```

## Global variables

```{r}
data_raw <- here::here(
  "data",
  "raw",
  "20260511_lemna_bacteria_coculture_frondarea"
)
data_processed <- here::here(
  "data",
  "processed",
  "04_lemna_bacteria_coculture_frondarea"
)

fs::dir_create(data_processed)
```

# Read and combine

Each xlsx file holds one `Export` sheet covering a subset of plates at a single timepoint. The day (0, 4, or 8) is encoded in the filename (e.g. `DW_plates_1-10_4d.xlsx`), not in the sheet itself, so we parse it from the filename as we read.

```{r}
#| output: true
frondarea_files <- fs::dir_ls(data_raw, glob = "*.xlsx")

read_frondarea_file <- function(path) {
  day <- as.integer(str_extract(path_file(path), "(?<=_)\\d+(?=d\\.xlsx$)"))

  read_excel(path, sheet = "Export") |>
    rename(
      date = Date,
      plate_id = `Plate ID`,
      plate_name = `Plate Name`,
      well_id = `Well ID`,
      sample = Sample,
      replicate = Replicate,
      area = Area,
      color_blue = `Average Object Color:Blue`,
      color_green = `Average Object Color:Green`,
      color_hue = `Average Object Color:Hue`,
      color_red = `Average Object Color:Red`,
      color_saturation = `Average Object Color:Saturation`,
      color_value = `Average Object Color:Value`,
      count = Count
    ) |>
    mutate(
      day = day,
      source_file = path_file(path)
    )
}

frondarea_raw <- frondarea_files |>
  map(read_frondarea_file) |>
  list_rbind()

glimpse(frondarea_raw)
```

# Derive plate and well identifiers

`plate_id` (e.g. `"A1"`) is **not** a globally unique plate identifier - it is reused across the different filename batches (plates 1-10, 11-14/15, 15-17), so the same code can refer to different physical plates. `plate_name` (the underscore-joined list of strains on that plate, e.g. `"HA_36_HA40_HA44"`) is unique per physical plate and - I verified this by comparing well-to-sample maps across files - stays constant across a plate's day 0/4/8 files when its `plate_id` changes between timepoints (e.g. the `SJ08_CONTROL` plate is `plate_id` `"A2"` at day 0 but `"E2"` at day 4/8). So `plate_name` alone is the correct cross-day plate key, with one exception: two distinct plates in the day-0-only file `DW_plates_15-17_0d.xlsx` both carry the generic `plate_name` `"CONTROL"` (`plate_id` `B2` and `C2`); for that one label we fall back to `plate_name` + `plate_id` to disambiguate them (they have no day 4/8 counterpart to match against, so this doesn't affect any longitudinal join). Combining `plate_uid` with `well_id` gives a stable identifier for a single physical well tracked across timepoints.

```{r}
#| output: true
frondarea <- frondarea_raw |>
  mutate(
    plate_uid = if_else(
      plate_name == "CONTROL",
      paste(plate_name, plate_id, sep = "_"),
      plate_name
    ),
    well_uid = paste(plate_uid, well_id, sep = "_"),
    strain = factor(sample) |> fct_relevel("CONTROL"),
    is_control = sample == "CONTROL"
  ) |>
  select(
    day,
    plate_uid,
    well_uid,
    well_id,
    plate_id,
    plate_name,
    strain,
    sample,
    replicate,
    is_control,
    area,
    count,
    color_blue,
    color_green,
    color_hue,
    color_red,
    color_saturation,
    color_value,
    date,
    source_file
  )

glimpse(frondarea)
```

# Quality checks

Each well should be measured at most once per day.

```{r}
#| output: true
# table should be empty
frondarea |>
  count(well_uid, day) |>
  filter(n > 1)
```

Area should be strictly positive - a few wells have `area == 0` and `count == 0` together. These are dropped

```{r}
#| output: true
zero_area <- frondarea |> filter(area <= 0)
zero_area |> select(day, plate_uid, well_id, sample, area, count, source_file)

frondarea <- frondarea |> filter(area > 0)
```

Sample-to-day completeness. Every strain should have (roughly) balanced replication across the three timepoints; `CONTROL` is expected to be imbalanced because a couple of `CONTROL`-only plates were only imaged at day 0 (see below).

```{r}
#| output: true
frondarea |>
  count(strain, day) |>
  pivot_wider(
    names_from = day,
    values_from = n,
    values_fill = 0,
    names_prefix = "day"
  ) |>
  arrange(strain)
```

`well_id` -> `sample` mapping should be identical across days for the same plate (confirms the longitudinal, same-well-tracked-over-time design rather than fresh plates each day).

```{r}
#| output: true
well_sample_by_day <- frondarea |>
  distinct(plate_uid, well_id, day, sample) |>
  pivot_wider(names_from = day, values_from = sample, names_prefix = "day")

# table should be empty
well_sample_by_day |>
  filter(if_any(starts_with("day"), \(x) {
    !is.na(x) & x != coalesce(day0, day4, day8)
  }))
```

`CONTROL` replication by plate and day, for context: a couple of pure-`CONTROL` plates (plate names `CONTROL` and `SJ08_CONTROL`, the latter only partially) were imaged only at day 0 and so contribute no day 4/8 growth information; the rest of `CONTROL` (plates `TL64_TL58_CONTROL` and `SJ08_CONTROL` at day 4/8) is measured at all three timepoints.

```{r}
#| output: true
frondarea |>
  filter(is_control) |>
  count(plate_uid, day)
```

# Exploratory look

```{r}
#| label: fig-frondarea-trajectories
#| output: true
#| fig-width: 10
#| fig-height: 6
#| fig-cap: "Per-well frond area trajectories by strain across the three imaging timepoints (day 0, 4, 8). CONTROL wells (no bacteria) are highlighted in red."
frondarea |>
  ggplot(aes(x = day, y = area, group = well_uid, color = is_control)) +
  geom_line(alpha = 0.3) +
  geom_point(size = 0.8, alpha = 0.5) +
  facet_wrap(~strain) +
  scale_color_manual(
    values = c(`TRUE` = "firebrick", `FALSE` = "grey40"),
    guide = "none"
  ) +
  labs(
    x = "Day",
    y = expression("Frond area (" * cm^2 * ")"),
    title = "Per-well frond area trajectories by strain"
  ) +
  theme_bw(base_size = 9)
```

# Write processed data

```{r}
write_rds(frondarea, path(data_processed, "frondarea.rds"))
write_tsv(frondarea, path(data_processed, "frondarea.tsv"))
```
