---
title: "Clustering/phylogenetics of 16S rRNA sequences from duckweed SynCom"
subtitle: ""
author: "Shane Hogle"
date: today
abstract: "Various operations to group the isolates into clusters using clustering algorithms and phylogenetic inference. At the end a phylogenetic tree for the isolates is plotting with associated taxonomic information as well as clustering identity based on shared kmers."
---

# Setup

Loads required libraries

```{r}
#| output: false
options(repos = c(CRAN = "https://packagemanager.posit.co/cran/__linux__/jammy/latest"))
library(tidyverse)
library(here)
library(fs)
library(DECIPHER)
```

# Read and format

Read 16S rRNA fasta sequences from the prior step.

```{r}
syncom16s_fna <- here::here("data", "sanger_seq", "16S_rRNA_SynCom.fna")
```

Convert to a DNA stringset object

```{r}
syncom16s_seqs <- readDNAStringSet(syncom16s_fna)
```

# Redundancy

All sequences are unique which is expected

```{r}
duplicated(syncom16s_seqs)
```

# Cluster

```{r}
set.seed(36742)
syncom16s_clusters <- Clusterize(syncom16s_seqs,
                       cutoff=0.01,     # at ~ species level (ie 99% identity)
                       minCoverage=0.5, # > 50% coverage
                       processors=NULL) # use all CPUs
# reset random seed
rm(.Random.seed, envir=globalenv())
```

# Classify (IDTAXA)

Using
[IDTAXA](https://microbiomejournal.biomedcentral.com/articles/10.1186/s40168-018-0521-5)
following the [approach shown
here](https://www2.decipher.codes/ClassifyOrganisms.html).

```{r}
#| eval: false
#| include: true
set.seed(234676)

# load the training data set
load(here::here("_data_raw", "idtaxa", "GTDB_r226-mod_April2025.RData"))

# classify the sequences
syncom16s_ids <- IdTaxa(syncom16s_seqs,
   trainingSet,
   strand="both", # or "top" if same as trainingSet
   threshold=60, # 60 (cautious) or 50 (sensible)
   processors=NULL) # use all available processors
# reset random seed
rm(.Random.seed, envir=globalenv())
```

```{r}
#| eval: false
#| include: false
write_rds(syncom16s_ids, here::here("data", "sanger_seq", "syncom16s_ids.rds"))
```

```{r}
#| eval: true
#| include: false
syncom16s_ids <- read_rds(here::here("data", "sanger_seq", "syncom16s_ids.rds"))
```

```{r}
syncom16s_idtx <- imap(
  syncom16s_ids,
  \(x, idx) tibble(a = x$rank, b = paste(x$taxon," (",round(x$confidence, digits = 1),"%)", sep = "")) %>%
    pivot_wider(names_from = a, values_from = b) %>%
    mutate(id = idx)
) %>%
  list_rbind() %>% 
  relocate(id)
```

# Classify (Blastn)

These steps need to be run outside of R/Rstudio

First we need to
[download](https://ftp.ncbi.nlm.nih.gov/blast/executables/LATEST/) and
[install](https://www.ncbi.nlm.nih.gov/books/NBK52640/) NCBI blast+
suite (I am using v2.16.0) and
[download](https://ftp.ncbi.nlm.nih.gov/blast/db/) the 16S_ribosomal_RNA
database.

```
blastn -num_threads 12 -db 16S_ribosomal_RNA -query 16S_rRNA_SynCom.fna -dust no -max_target_seqs 5 -out 16S_rRNA_SynCom_v_blast16SDB.tsv -outfmt "6 qseqid sseqid pident length mismatch gapopen qstart qend sstart send evalue bitscore staxid"
```

The last column contains the taxonomy ID from the NCBI 16S Ribosomal
database. We will just take the taxonomic lineage of the best hit

```
csvtk tab2csv 16S_rRNA_SynCom_v_blast16SDB.tsv | csvtk cut -f1,13 | csvtk spread -k1 -v2 | sed 's/^,//g' | csvtk transpose | cut -d ";" -f1 | csvtk cut -f2,1 | csvtk csv2tab > a
csvtk tab2csv 16S_rRNA_SynCom_v_blast16SDB.tsv | csvtk cut -f1,13 | csvtk spread -k1 -v2 | sed 's/^,//g' | csvtk transpose | cut -d ";" -f1 | csvtk cut -f2 | taxonkit lineage -n -L > b
# combine into one file
paste a b > blastn_lineage_best_hit.tsv
rm a b
```

Now read this file back into R

```{r}
#| echo: true
#| output: false
syncom16s_blastn <- readr::read_tsv(here::here("data", "sanger_seq", "blastn_lineage_best_hit.tsv"),
                col_names = FALSE) %>% 
  dplyr::select(id = X2, ncbi_taxid = X1, ncbi_best_species = X4)
```

# Combine clustering and classification

```{r}
syncom16s_idtx_lineage_info <- syncom16s_idtx %>% 
  left_join(rownames_to_column(syncom16s_clusters, var = "id"), by = join_by(id)) %>% 
  left_join(syncom16s_blastn, by = join_by(id)) %>% 
  relocate(ncbi_best_species, .after="genus") %>% 
  relocate(cluster, .after="id") %>% 
  arrange(cluster)
```

# 16S rRNA sequence alignment

We start with DNA sequences and convert to RNA so that DECIPHER can leverage structural information from the RNA

```{r}
syncom16s_seqs_rna <- RNAStringSet(syncom16s_seqs)
```

Align the sequences

From [DECIPHER documentation on sequence
alignment](https://bioconductor.org/packages/devel/bioc/vignettes/DECIPHER/inst/doc/ArtOfAlignmentInR.pdf)

> 5.2 Example: Non-coding RNA sequences Much like proteins, non-coding
> RNAs often have a conserved secondary structure that can be used to
> improve their alignment. The PredictDBN function will predict base
> pairings from a sequence alignment by calculating the mutual
> information between pairs of positions. If RNA sequences are given as
> input, AlignSeqs will automatically use the output of PredictDBN to
> iteratively improve the alignment. Providing an RNAStringSet also
> causes single-base and double-base substitution matrices to be used,
> and is preferable to providing a DNAStringSet when the sequences are
> non-coding RNA. The type of the input sequences can easily be
> converted to RNA, as shown below.

```{r}
#| eval: false
#| include: true
set.seed(2342)
syncom16s_seqs_rna_aligned <- AlignSeqs(syncom16s_seqs_rna)
# reset random seed
rm(.Random.seed, envir=globalenv())
```

```{r}
#| eval: false
#| include: false
write_rds(syncom16s_seqs_rna_aligned, here::here("data", "sanger_seq", "syncom16s_seqs_rna_aligned.rds"))
```

```{r}
#| eval: true
#| include: false
syncom16s_seqs_rna_aligned <- read_rds(here::here("data", "sanger_seq", "syncom16s_seqs_rna_aligned.rds"))
```

```{r}
# Write as aligned fasta DNA file 
writeXStringSet(DNAStringSet(syncom16s_seqs_rna_aligned),
                here::here("data", "sanger_seq", "16S_rRNA_SynCom_aligned.fna"),
                format='fasta')
```

# Phylogenetic tree building

Following maximum likelihood tutorial [DECIPHER documentation on
phylogenetic
trees](https://bioconductor.org/packages/devel/bioc/vignettes/DECIPHER/inst/doc/GrowingTrees.pdf)

This step constructs a phylogenetic tree using the general time reversable (GTR) model with a discrete Gamma model [(Yang, 1994)](https://link.springer.com/article/10.1007/BF00160154) with the default 4 rate categories. Branch support is depicted using aBayes probabilities, which is a Bayesian-like transformation of an approximate Likelihood Ratio Test [(Anisimova et al., 2011)](https://academic.oup.com/sysbio/article/60/5/685/1644562). 

```{r}
#| eval: false
#| include: true
set.seed(26345)
syncom16s_seqs_mltree <- Treeline(
  syncom16s_seqs_rna_aligned,
  method = "ML",
  model = "GTR+G4",
  verbose = TRUE,
  processors = 12
)
# reset random seed
rm(.Random.seed, envir=globalenv())
```

```{r}
#| eval: false
#| include: false
write_rds(syncom16s_seqs_mltree, here::here("data", "sanger_seq", "syncom16s_seqs_mltree.rds"))
```

```{r}
#| eval: true
#| include: false
syncom16s_seqs_mltree <- read_rds(here::here("data", "sanger_seq", "syncom16s_seqs_mltree.rds"))
```

```{r}
# this function adds an 'edgetext' attribute that is recognized by WriteDendrogram so that
# branch support values (aBayes) are written properly
add_boots <- function(node){
  s <- attr(node, "probability")
  if (!is.null(s))
    attr(node, "edgetext") <- formatC(as.numeric(s), digits=2, format="f")
  return(node)
}

syncom16s_seqs_mltree <- dendrapply(syncom16s_seqs_mltree, add_boots)
```

Save ML tree as newick file

```{r}
WriteDendrogram(syncom16s_seqs_mltree, 
                file=here::here("data", "sanger_seq", "syncom16s_seqs_mltree.nwk"),
                # quote ="" ensure that bootstraps are not written quoted as strings
                # so that ggtree doesn't fail to parse them
                quote="")
```

Tree building results

```         
Fitting initial tree to model:
GTR+G4 -ln(L) = 15520, AICc = 31633, BIC = 32781

Optimizing up to 400 candidate trees:
Tree #136. -ln(L) = 15395.993 (0.000%), 9 Climbs, 0 Grafts of 7                                                                                                       

Finalizing the best tree (#130):
-ln(L) = 15395.993 (0.000%), 1 Climb                                                                                                                                  

Model parameters:
Frequency(A) = 0.244
Frequency(C) = 0.223
Frequency(G) = 0.312
Frequency(T) = 0.221
Rate A <-> C = 0.816
Rate A <-> G = 1.978
Rate A <-> T = 1.262
Rate C <-> G = 0.677
Rate C <-> T = 3.029
Rate G <-> T = 1.000
Alpha = 0.302

Time difference of 1090.3 secs
```

# Aggregate various taxa information

```{r}
#| echo: true
#| output: false
sanger_batch_info <- read_tsv(here::here("data", "sanger_seq", "sanger_batch_info_01.tsv"))
```

```{r}
final_table <- left_join(syncom16s_idtx_lineage_info, sanger_batch_info, by = join_by(id == strainID)) %>% 
  dplyr::rename(strainID = id, taxa_cluster = cluster) %>% 
  mutate(genus1 = stringr::str_extract(genus, "(^[:alpha:]*).*", group = 1)) %>% 
  mutate(genus1 = case_when(is.na(genus1) | genus1 == "unclassified" ~ stringr::str_extract(ncbi_best_species, "^[:alpha:]*"),
                           TRUE ~ genus1)) %>% 
  mutate(tree_name = paste(genus1, strainID)) %>% 
  relocate(c(bacteria_plate_well, location, google_drive_id, tree_name), .after = strainID) %>% 
  relocate(google_drive_id, taxa_cluster, tree_name, .after = strainID) %>% 
  relocate(ncbi_best_species, ncbi_taxid, .after = location) %>% 
  dplyr::select(-genus1)

write_tsv(final_table, here::here("data", "strain_table_info.tsv"))
```
