---
title: "Bacterial isolate taxonomy from SILVA ACT classification"
subtitle: ""
author: "Shane Hogle"
date: today
abstract: "Parses SILVA ACT classifications of the consensus 16S sequences assembled in `01_sanger_seq/01_trace_qc_assembly.qmd`, builds a per-isolate taxonomy table, and summarizes genus/family/phylum composition and phylogeny."
---

The consensus 16S sequences produced in [01.A Trace QC + assembly](../01_sanger_seq/01_trace_qc_assembly.qmd) were submitted to the [SILVA ACT web aligner](https://www.arb-silva.de/aligner) with LCA classification against SILVA SSU and GTDB enabled. This notebook reads back the two downloaded ACT outputs — the classification result list and the guide tree — and joins them to the per-isolate assembly metadata to build the taxonomy table and summary figures.

# Setup

## Libraries

Loads required libraries

```{r}
library(tidyverse)
library(here)
library(fs)
library(ape)
library(treeio)
library(ggtree)
library(RColorBrewer)
library(Polychrome)
library(DT)

source(here::here("R", "plotting.R"))
```

## Global variables

```{r}
data_raw <- here::here("data", "raw", "20260908_silva_act")
data_processed_sanger <- here::here("data", "processed", "01_sanger_seq")
data_processed <- here::here("data", "processed", "02_bac_taxonomy")

fs::dir_create(data_processed)

silva_resultlist <- here::here(
  data_raw,
  "arb-silva.de_align_resultlist_1508237.csv"
)
silva_tree_file <- here::here(
  data_raw,
  "arb-silva.de_2026-09-08_id1508237.tree"
)

outgroup_genus <- "Lactococcus"
```

# Read and parse SILVA ACT output

SILVA SSU lineage strings are semicolon-delimited as `Domain;<unnamed>;Phylum;Class;Order;Family;Genus`, with the second field always empty, and ambiguous LCA nodes suffixed with `--other`. GTDB lineage strings instead carry `g__`/`s__` prefixed tokens. The two helpers below pull out the ranks we need from each scheme.

```{r}
parse_silva_taxonomy <- function(tax_string) {
  ranks <- c("domain", "phylum", "class", "order", "family", "genus")
  rank_idx <- c(1, 3, 4, 5, 6, 7)
  empty <- c("None", "Unclassified;", "")

  if (is.na(tax_string) || tax_string %in% empty) {
    return(
      set_names(as.list(rep(NA_character_, length(ranks))), ranks) |>
        as_tibble()
    )
  }

  parts <- tax_string |>
    str_remove(";$") |>
    str_split(";") |>
    pluck(1) |>
    str_trim()
  parts <- parts[parts != ""]
  parts <- str_remove(parts, "--other")

  vals <- map_chr(
    rank_idx,
    ~ if (length(parts) >= .x) parts[.x] else NA_character_
  )
  set_names(as.list(vals), ranks) |> as_tibble()
}

parse_gtdb_taxonomy <- function(tax_string) {
  empty <- c("None", "Unclassified;", "")
  if (is.na(tax_string) || tax_string %in% empty) {
    return(tibble(gtdb_genus = NA_character_, gtdb_species = NA_character_))
  }

  parts <- tax_string |>
    str_remove(";$") |>
    str_split(";") |>
    pluck(1) |>
    str_trim()
  parts <- parts[parts != ""]

  tibble(
    gtdb_genus = parts |>
      str_subset("^g__") |>
      str_remove("^g__") |>
      pluck(1, .default = NA_character_),
    gtdb_species = parts |>
      str_subset("^s__") |>
      str_remove("^s__") |>
      pluck(1, .default = NA_character_)
  )
}
```

Read the result list and apply the parsers row-wise

```{r}
silva_raw <- read_delim(
  silva_resultlist,
  delim = ";",
  quote = '"',
  show_col_types = FALSE
)

silva_tax <- silva_raw$lca_tax_slv |> map_dfr(parse_silva_taxonomy)
gtdb_tax <- silva_raw$lca_tax_gtdb |> map_dfr(parse_gtdb_taxonomy)

silva_parsed <- silva_raw |>
  select(
    seq_id = sequence_identifier,
    silva_identity = identity,
    silva_quality = quality,
    silva_score = sequence_score
  ) |>
  bind_cols(silva_tax, gtdb_tax)
```

# Build taxonomy table

Join the SILVA classification onto the per-isolate assembly metadata from the trace QC/assembly step

```{r}
assemblies <- read_csv(
  here::here(data_processed_sanger, "assemblies.csv"),
  show_col_types = FALSE
)

taxonomy_table <- assemblies |>
  select(
    seq_id,
    batch,
    well,
    bacteria_plate_well,
    location,
    google_drive_id,
    assembly_type,
    consensus_length,
    fwd_trimmed_length,
    rev_trimmed_length,
    overlap_length,
    fwd_rev_overlap_identity = overlap_pct_identity
  ) |>
  left_join(silva_parsed, by = "seq_id")

write_csv(
  taxonomy_table,
  here::here(data_processed, "isolate_taxonomy_table.csv")
)
```

```{r}
#| output: true
n_genus <- sum(!is.na(taxonomy_table$genus))
n_genera <- n_distinct(na.omit(taxonomy_table$genus))
n_unclassified <- sum(is.na(taxonomy_table$domain))
n_gtdb_species <- sum(!is.na(taxonomy_table$gtdb_species))

tibble(
  metric = c(
    "isolates",
    "genus-level calls",
    "distinct genera",
    "unclassified",
    "GTDB species calls"
  ),
  value = c(
    nrow(taxonomy_table),
    n_genus,
    n_genera,
    n_unclassified,
    n_gtdb_species
  )
)
```

Below is the compiled taxonomy table: SILVA genotype/quality metrics from the ACT alignment plus `domain` through `genus` from the SILVA SSU LCA classification, and `gtdb_genus`/`gtdb_species` from the GTDB LCA classification.

```{r}
#| output: true
DT::datatable(
  taxonomy_table,
  class = "cell-border stripe",
  rownames = FALSE,
  extensions = "FixedColumns",
  options = list(
    dom = "ft",
    pageLength = -1,
    scrollX = TRUE,
    fixedColumns = TRUE
  )
)
```

# Taxonomy figures

The panels below summarize isolate composition and phylogeny. Each is rendered as its own figure rather than as a combined multi-panel layout; a separate Sanger trace QC figure already lives in the [01.A Trace QC + assembly](../01_sanger_seq/01_trace_qc_assembly.qmd) notebook and isn't repeated here.

Only isolates with a resolved SILVA SSU classification (`domain` not missing) are used below

```{r}
classified <- taxonomy_table |> filter(!is.na(domain))
```

A shared palette for phylum, plus a helper that bins a taxonomic rank into its top categories (grouping everything else as `"Other"`) and computes relative composition per sampling site. Plot styling (`theme_project()`, `capit()`) and the `make_palette()` generator live in [`R/plotting.R`](../../R/plotting.R), shared with other notebooks in this project.

```{r}
phylum_colors <- c(
  "Pseudomonadota" = "#1f78b4",
  "Bacteroidota" = "#ff7f00",
  "Bacillota" = "#33a02c",
  "Deinococcota" = "#e31a1c"
)

composition_by_site <- function(
  df,
  rank_col,
  top_n,
  palette,
  other_color = "grey60"
) {
  top_cats <- df |>
    filter(!is.na(.data[[rank_col]])) |>
    count(.data[[rank_col]], sort = TRUE) |>
    slice_head(n = top_n) |>
    pull(1)

  plot_df <- df |>
    filter(!is.na(location)) |>
    mutate(
      cat = if_else(.data[[rank_col]] %in% top_cats, .data[[rank_col]], "Other")
    ) |>
    count(location, cat) |>
    group_by(location) |>
    mutate(pct = 100 * n / sum(n)) |>
    ungroup() |>
    mutate(cat = factor(cat, levels = c(top_cats, "Other")))

  cat_colors <- c(palette[top_cats], "Other" = other_color)

  ggplot(plot_df, aes(x = location, y = pct, fill = cat)) +
    geom_col(color = "white", linewidth = 0.3) +
    scale_fill_manual(values = cat_colors, name = str_to_title(rank_col)) +
    labs(x = NULL, y = "% of isolates") +
    capit() +
    theme_project() +
    theme(
      legend.position = "right",
      legend.title = element_text(),
      axis.text.x = element_text(angle = 20, hjust = 1)
    )
}
```

## Genus composition across all isolates

```{r}
genus_counts <- classified |> count(genus, sort = TRUE)

top_genus <- genus_counts |> filter(!is.na(genus)) |> slice_head(n = 12)
other_n <- genus_counts |>
  filter(!genus %in% top_genus$genus | is.na(genus)) |>
  summarise(n = sum(n)) |>
  pull(n)

genus_plot_df <- bind_rows(top_genus, tibble(genus = "Other", n = other_n)) |>
  mutate(genus = fct_reorder(genus, n))
```

```{r}
#| label: fig-genus-counts
#| output: true
#| fig-width: 7
#| fig-height: 6
#| fig-cap: !expr paste0("Genus composition across all ", nrow(classified), ' classified isolates. The 12 most abundant genera are shown individually; all remaining genera are pooled as "Other".')
ggplot(genus_plot_df, aes(x = n, y = genus)) +
  geom_col(fill = "#1f78b4") +
  geom_text(aes(label = n), hjust = -0.3, size = 3) +
  labs(x = "Number of isolates", y = NULL) +
  scale_x_continuous(expand = expansion(mult = c(0, 0.1))) +
  capit() +
  theme_project()
```

## Isolates per sampling site

```{r}
site_counts <- classified |>
  mutate(location = replace_na(location, "No site info")) |>
  count(location)
```

```{r}
#| label: fig-site-counts
#| output: true
#| fig-width: 6
#| fig-height: 5
#| fig-cap: "Number of classified isolates recovered per sampling site. Isolates lacking plate-map location metadata are grouped as \"No site info\"."
ggplot(site_counts, aes(x = location, y = n)) +
  geom_col(fill = "#1b9e77") +
  geom_text(aes(label = n), vjust = -0.4, size = 3.5, fontface = "bold") +
  labs(x = NULL, y = "Number of isolates") +
  scale_y_continuous(expand = expansion(mult = c(0, 0.12))) +
  capit() +
  theme_project()
```

## Relative composition by site — phylum

```{r}
#| label: fig-phylum-composition
#| output: true
#| fig-width: 6
#| fig-height: 5
#| fig-cap: "Relative phylum composition of classified isolates by sampling site."
composition_by_site(classified, "phylum", top_n = 10, palette = phylum_colors)
```

## Relative composition by site — family

```{r}
#| label: fig-family-composition
#| output: true
#| fig-width: 8
#| fig-height: 5
#| fig-cap: "Relative family composition of classified isolates by sampling site. The 15 most abundant families are shown individually; all remaining families are pooled as \"Other\"."
family_cats <- classified |>
  filter(!is.na(family)) |>
  distinct(family) |>
  pull(family)
composition_by_site(
  classified,
  "family",
  top_n = 15,
  palette = make_palette(family_cats)
)
```

## Relative composition by site — genus

```{r}
#| label: fig-genus-composition
#| output: true
#| fig-width: 8
#| fig-height: 5
#| fig-cap: "Relative genus composition of classified isolates by sampling site. The 15 most abundant genera are shown individually; all remaining genera are pooled as \"Other\"."
genus_cats <- classified |>
  filter(!is.na(genus)) |>
  distinct(genus) |>
  pull(genus)
composition_by_site(
  classified,
  "genus",
  top_n = 15,
  palette = make_palette(genus_cats)
)
```

## Phylogeny

The SILVA ACT guide tree places every isolate among several hundred SILVA reference sequences, which is useful for classification but too dense to read as a figure. For the phylogeny below we root the guide tree on an isolate classified as *Lactococcus* (the most basal genus recovered, Bacillota) and then prune it down to just our isolates, keeping the topology induced by the full guide tree alignment.

```{r}
tree <- ape::read.tree(silva_tree_file)

outgroup_tip <- classified |>
  filter(genus == outgroup_genus, seq_id %in% tree$tip.label) |>
  pull(seq_id) |>
  pluck(1)

tree_rooted <- ape::root(tree, outgroup = outgroup_tip, resolve.root = TRUE)

isolate_tips <- intersect(tree_rooted$tip.label, classified$seq_id)
tree_pruned <- ape::keep.tip(tree_rooted, isolate_tips)
```

Every genus recovered among the tree's isolates gets its own tip color below — nothing is pooled into an "Other" category. Class-level groups are additionally marked with a `ggtree::geom_cladelab()` bracket. `geom_cladelab()` needs a single tree node spanning exactly the tips being labeled, but because this is a guide tree built for alignment across hundreds of SILVA references rather than a phylogeny inferred from just our isolates, a class doesn't always come out as a clean monophyletic clade after pruning. The helper below finds, for a given set of tips, the largest node whose full descendant set is "pure" (i.e. contains only members of that class and nothing else) — the clade that gets labeled — and reports how many of the class's tips that clade actually covers.

```{r}
node_tip_set <- function(tree, node) {
  ntip <- length(tree$tip.label)
  if (node <= ntip) {
    return(tree$tip.label[node])
  }
  ape::extract.clade(tree, node)$tip.label
}

largest_pure_clade <- function(tree, target_tips) {
  n_total <- length(target_tips)

  if (n_total == 1) {
    return(tibble(
      node = which(tree$tip.label == target_tips),
      n_covered = 1L,
      n_total = 1L
    ))
  }

  ntip <- length(tree$tip.label)
  internal_nodes <- seq(ntip + 1, ntip + tree$Nnode)
  sizes <- map_int(internal_nodes, function(nd) {
    desc <- node_tip_set(tree, nd)
    if (all(desc %in% target_tips)) length(desc) else NA_integer_
  })

  if (all(is.na(sizes))) {
    return(tibble(node = NA_integer_, n_covered = 0L, n_total = n_total))
  }

  best_idx <- which.max(sizes)
  tibble(
    node = internal_nodes[best_idx],
    n_covered = sizes[best_idx],
    n_total = n_total
  )
}

class_groups <- classified |>
  filter(seq_id %in% isolate_tips, !is.na(class)) |>
  summarise(members = list(seq_id), .by = class) |>
  mutate(clade = map(members, ~ largest_pure_clade(tree_pruned, .x))) |>
  unnest(clade) |>
  select(-members) |>
  filter(!is.na(node))
```

```{r}
n_class_partial <- sum(class_groups$n_covered != class_groups$n_total)
```

All `r nrow(class_groups)` classes recovered are labeled; `r n_class_partial` of them (*Bacilli*, whose *Lactococcus* and *Exiguobacterium* members don't share a single common ancestor exclusive of other classes here) is only partially monophyletic, so its bracket covers just the largest pure block.

```{r}
all_genus_tree <- classified |>
  filter(seq_id %in% isolate_tips, !is.na(genus)) |>
  distinct(genus) |>
  arrange(genus) |>
  pull(genus)

genus_tree_pal <- make_palette(all_genus_tree)

tip_data <- classified |>
  filter(seq_id %in% isolate_tips) |>
  transmute(label = seq_id, genus_grp = coalesce(genus, "Unclassified"))
```

```{r}
#| label: fig-tree
#| output: true
#| fig-width: 11
#| fig-height: 22
#| fig-cap: !expr paste0("Phylogeny of the ", length(isolate_tips), " classified isolates, pruned from the SILVA ACT guide tree and rooted on an isolate classified as *Lactococcus*. Tip points are colored by genus, with every genus recovered (", length(all_genus_tree), ') shown individually; isolates with a domain-level SILVA classification but no resolved genus are labeled "Unclassified". Bracketed labels on the right mark class-level membership (`ggtree::geom_cladelab()`). Scale bar represents substitutions per site.')
ggtree(tree_pruned, layout = "rectangular") %<+%
  tip_data +
  geom_tippoint(aes(color = genus_grp), size = 1.5) +
  geom_tiplab(size = 2, hjust = -0.1) +
  scale_color_manual(
    values = c(genus_tree_pal, "Unclassified" = "grey50"),
    name = "Genus"
  ) +
  geom_cladelab(
    node = class_groups$node,
    label = class_groups$class,
    align = TRUE,
    geom = "text",
    offset = 0.9,
    offset.text = 0.05,
    fontsize = 3,
    fontface = "bold",
    barsize = 0.6
  ) +
  guides(color = guide_legend(ncol = 1, override.aes = list(size = 3))) +
  ggplot2::xlim(NA, 4) +
  geom_treescale(offset = 1)
```
