---
title: "Read and format Sanger sequences from Duckweed"
subtitle: ""
author: "Shane Hogle"
date: today
abstract: "This reads Fwd/Rev Sanger sequencing reads of duckweed marker, trims them, overlaps them, and produces fasta condtigs for use in downstream analysises."
---

We wanted to know whether the duckweeds that we isolated bacteria from are L. minor. To determine species identity we followed the protocol from the publication:

> Borisjuk N, Chu P, Gutierrez R, Zhang H, Acosta K, Friesen N, Sree KS, Garcia C, Appenroth KJ, Lam E. 2015. Assessment, validation and deployment strategy of a two-barcode protocol for facile genotyping of duckweed species. Plant Biol  17 Suppl 1:42–49. <https://onlinelibrary.wiley.com/doi/10.1111/plb.12229>
  
This protocol also references: 

> CBOL Plant Working Group1,P.M. Hollingsworth,L.L. Forrest,J.L. Spouge,M. Hajibabaei,S. Ratnasingham,M. van der Bank,M.W. Chase,R.S. Cowan,D.L. Erickson,A.J. Fazekas,S.W. Graham,K.E. James,K. Kim,W.J. Kress,H. Schneider,J. van AlphenStahl,S.C. Barrett,C. van den Berg,[...] & D.P. Little,  A DNA barcode for land plants, Proc. Natl. Acad. Sci. U.S.A. 106 (31) 12794-12797 (2009). <https://doi.org/10.1073/pnas.0905845106>

> Wang, W., Wu, Y., Yan, Y. et al. DNA barcoding of the Lemnaceae, a family of aquatic monocots. BMC Plant Biol 10, 205 (2010). <https://doi.org/10.1186/1471-2229-10-205>

Briefly, we PCR amplify and sanger sequence two plastid barcoding regions *atpF-atpH* and *psbK-psbI* to assign species identity.


## Primers and PCR optimum annealing:

| Marker region | Primer sequence Forward     | Primer sequence Reverse        | Amplicon size (Lemna minor) | Ta Optimum (Lemna minor) |
|---------------|-----------------------------|--------------------------------|-----------------------------|--------------------------|
| psbK-psbI     | 5′-TTAGCATTTGTTTGGCAAG-3′   | 5′- AAAGTTTGAGAGTAAGCAT-3′     | 544 bp                      | 51°C                     |
| atpF-atpH     | 5′-ACTCGCACACACTCCCTTTCC-3′ | 5′- GCTTTTATGGAAGCTTTAACAAT-3′ | 675 bp                      | 53°C                     |

## Protocol:
Briefly, 50-100 ng genomic DNA and 5 pmol of each primer are added with the JumpStart™Redtop® ReadyMix™Reaction Mix (P1107, Sigma) Redix in 25 ul of final volume.

Following amplification, the DNA fragments were purified using the ExoSAP-IT kit (Affymetrix, Santa Clara, CA, USA) and for Sanger sequencing.

## Libraries

```{r}
#| eval: false
#| include: true
library(tidyverse)
library(here)
library(fs)
library(sangeranalyseR)
library(Biostrings)
```

## Global variables

```{r}
#| eval: false
#| include: true
data_raw <- here::here("data", "raw", "20251105_duckweed_genotypeing")
data_processed <- here::here("data", "processed", "03_duckweed_genotyping")

fs::dir_create(data_processed)
```

# Read and format

This code follows the [instructions here](https://sangeranalyser.readthedocs.io/en/latest/content/quickstart.html). Basically, the function reads Fwd/Rev traces trims them using "M1 - Modified Mott Trimming" algorithm, overlaps them then aligns them

```{r}
#| eval: false
#| include: true
contigs_atpFH <- sangeranalyseR::SangerAlignment(
  ABIF_Directory = here::here(data_raw, "renamed_ab1_files", "atpFH"),
  # we are being a bit permissive with the read qualities here since the Sanger traces
  # weren't super good. But they are good enough for assigning species identity
  minReadLength = 1,
  minReadsNum = 1,
  M1TrimmingCutoff = 0.05,
  REGEX_SuffixForward = "_fwd.ab1$",
  REGEX_SuffixReverse = "_rev.ab1$"
)

contigs_atpFH@objectResults@readResultTable

contigs_psbKI <- sangeranalyseR::SangerAlignment(
  ABIF_Directory = here::here(data_raw, "renamed_ab1_files", "psbKI"),
  # we are being a bit permissive with the read qualities here since the Sanger traces
  # weren't super good. But they are good enough for assigning species identity
  minReadLength = 1,
  minReadsNum = 1,
  M1TrimmingCutoff = 0.05,
  REGEX_SuffixForward = "_fwd.ab1$",
  REGEX_SuffixReverse = "_rev.ab1$"
)

contigs_psbKI@objectResults@readResultTable
```

### Write fasta

```{r}
#| eval: false
#| include: true
sangeranalyseR::writeFasta(contigs_atpFH, data_processed)

sangeranalyseR::writeFasta(contigs_psbKI, data_processed)
```

```{r}
#| eval: false
#| include: false
# saving intermediate files
write_rds(contigs_atpFH, here::here(data_processed, "atpFH.rds"))

write_rds(contigs_psbKI, here::here(data_processed, "psbKI.rds"))
```
